---
title: "Relationship Between Unemployment Rates Between Ages 25-34 From 2008 to 2025"
author: "Kei Matsumoto"
thanks: "Project repository available at: [https://github.com/keimatsumoto1/261A](https://github.com/keimatsumoto1/261A)"
date: 9/22/2025
date-format: long
abstract: "Employment through the years always changes with trends of industries as well as international events/disasters. However, with the increase in possible startup opportunities through new discoveries like AI, has unemployment rates changed over the years? A dataset provided by the California Open Data pool highlights the overall unemployment rates over a monthly period from 2008 to 2025 within specific age groups. This paper will cover the possibility to fit a simple linear regression model to see if there is a relation between unemployment rate and time for people between ages 25 and 34. Overall in this study, it is hard to find if the year correlates to the unemployment rates through linear regression as a different model is most likely better to analyze this data."
format: pdf
number-sections: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false
# load libraries
library(readr)
library(ggplot2)
unemployment <- read_csv("unemployment_rate_by_age_groups.csv")
```

# Introduction

Just by ear, there are people who say that there are better job
opportunities with the rise in new advancements in AI and others who say
that it is much more of a challenge than with the increase in
competitiveness for those positions. Most people will start their
careers after graduating from either undergraduate or graduate school
around the age of 25-34 which will be the age range that will be
analyzed.

This paper will only cover simple linear regression and linear modeling
with a lack of further analysis. In this paper, we will conclude that
linear regression may not be the optimal model to use to see the
relationship between time and unemployment rates as the data highlights
a possibility of a cyclic pattern due to sudden spikes of unemployment
in specific years. One large spike found was during 2019-2021 which can
be due to the breakout of Covid-19 with many jobs having in-person
interactions. The linear regression will be analyzed through the
programming language R using the basic functions within R and major
packages such as "lm" and "ggplot" to visualize the data.

The paper will start with the description of the data used, into the
methods conducted, and lastly the explanation of the results we have
possibly found through the use of the statistical analysis within our
knowledge of linear regression.

# Data {#sec-data}

```{r fig-}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot comparing the year of survey study as the x axis, and the rate of unemployment between 0 and 1 as the y axis with the fitted line of regression."


ggplot(unemployment, aes(x = Year, y =  `Age 25-34`))+
  geom_point() +
  geom_smooth(method = "lm") + 
  xlab("Year") + 
  ylab("Unemployment Rate (0-1)")+
  ggtitle("Scatterplot of Relationship Between Year and Unemployment Rate Between Ages 25-34")+
  theme(plot.title = element_text(size = 12))


```

This data was retrieved from the California Open Data Pool titled
"Unemployment Rate by Age Groups" where the Current Population Survey
(CPS) conducted a non-seasonally adjusted survey between 7 different age
groups for each month of 2008 to 2025. The data is retrieved from the
Employment Development Department who releases all of the employment
statistics such as civilian labor forces, unemployment rates, and
industry employment by geography.(CPS, 2025) The EDD collaborates with
the U.S. Bureau of Labor Statistics and the exact calculation of the
unemployment rate is defined by (unemployed citizens / civilian labor
force). Some months are missing in this dataset where January of 2008
and September through December of 2025 are not included. There does not
appear to be any missing data within the dataset besides the last months
of 2025. Therefore, the data for 2025 may not be as accurate as a
quarter of the year is missing compared to other years due to the date
of analysis is still being updated through each month.

The data uses measurements of percentages of unemployed people in the
designated age group from the survey with other columns being
demographic area, date, year, month, and each age group's rate of
unemployment.

# Methods {#sec-methods}

In this study we will be conducting a simple linear regression to
analyze the relationship between the rate of unemployment and time
specifically in years ranging from 2008 to 2015.

$$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$

$\beta_0$ will represent the intercept of the model which in this
analysis will be the expected unemployment rate during the first time
the study was conducted which is February of 2008.

$\beta_1$ will represent the slope of the model which in this analysis
will be the expected decrease in unemployment rate through each increase
of 1 year.

This model analysis will be conducted through the programming language R
(insert citation) with several packages such as ggplot and readr.

# Results {#sec-results}

```{r lm}
#| echo: false
#| warning: false
#| message: false
lm = lm(`Age 25-34`~Year, data = unemployment)

```

$$\beta_0 = 6.56$$ $$\beta_1 = -0.0032195$$ $$R^2 = 0.3847$$

Within our simple linear regression we found that our $\beta_0$is 6.56
and our $\beta_1$ is -0.0032195. Our $\beta_0$ in this scenario does not
apply to our analysis due to the unemployment rate not being able to
exceed 1. We can possibly say that our unemployment rate is close to 1
before 2008 which is due to the lack of studies made before 2008 in our
experiment. Our $\beta_1$ is slightly negative indicating a very minor
decrease over the years between 2008 and 2025. To be exact, for every
additional year that passes, the rate of unemployment will on average
decrease by 0.0032195 or 0.32195%.

In addition, our linear regression analysis includes and R-squared value
of 0.3847 indicating that the fit of a linear model may not be ideal for
this data.

```{r residual}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Residual plot of the x axis being the fitted values and the y axis being the residuals for a simple linear regression with year being the respone and unemployment rate being the predictor."
p <- data.frame(
  res = resid(lm),
  fit = fitted(lm)
)
ggplot(p, aes(x = fit, y = res)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Residual (observed - fitted)") +
  xlab("Fitted value") +
  ggtitle("Residuals vs. Fitted plot")


```

```{r qq}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "QQ plot between the sample (x) and theoretical (y) quantiles"

residual = residuals(lm)
df1 = data.frame(residuals = residual)

qqnorm(residual)
qqline(residual)

```

Our residuals vs fitted plots highlight that our model isn't the best
fit for a simple linear regression where the variation of points are not
in a form that shows no trends. With our data's sample size being 211
observations, we can possibly infer that our variances are equal.

Our qq plot on the other hand suggests that our model has a possibility
of having a good fit to our data. However, our tail and head of the QQ
plot highlights the variance from the fit which can add on to our
R-squared value being low.

# Discussion {#sec-discussion}

In our analysis, we tested to see if we can fit a simple linear
regression to find if unemployment rates have a relationship over time.
We discovered that 2010 and 2020 have large peeks and sharp rises of
unemployment rates which can be due to outside influences.

2008 was one of our spikes in unemployment rates. The Great Recession
which was around the time can be a possible issue identified by the
large spike in the beginning of our data. Another spike was identified
around 2020 which highlights possibilites of influence by COVID 19
affecting the unemployment rate drastically. A study from US Bureau of
Statistics indicates that in 2019 found that roughly 3/4 of the workers
in the United States had to work with in-person human interaction. (U.S.
Bureau of Labor Statistics, 2023) Combining this statistic with the lock
down protocol from COVID-19 can possibly connect to the drastic increase
in unemployment during 2020. By noticing that the decrease then
continues after the two large spikes, we can possibly see that the
unemployment rate possibly would have slowly decreased over time without
any interference.

Due to our R-squared value being considerably low and multiple
violations against our simple linear regression conditions such as
unequal variance and normality of errors, a better model can be possibly
used to fit this data. Possible indications of a cyclical or time-series
related model can be found from observing the scatter plot that can be
conducted in a later study.

# References

H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag
New York, 2016.

R Core Team (2023). *R: A Language and Environment for Statistical
Computing*. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>.

Unemployment rate by age groups - Unemployment rate by age groups -
California Open data. (n.d.).\
https://data.ca.gov/dataset/unemployment-rate-by-age-groups/resource/be49fea7-af13-4781-8113-4bb66ba508e9

U.S. Bureau of Labor Statistics. (n.d.). Three-fourths of workers had to
interact with the public in 2019; 4.3 percent worked around crowds. U.S.
Bureau of Labor Statistics.
https://www.bls.gov/opub/ted/2020/three-fourths-of-workers-had-to-interact-with-the-public-in-2019-4-3-percent-worked-around-c
rowds

# Things that I will need think about adding in the final draft

1.  Citation with bib file
2.  Citations within text
3.  Possibly using more than just ages 25-34 and add a better
    combination graph with deeper analysis (would this be beneficial or
    too much?)
4.  sectioning
