---
title: "Relationship Between Unemployment Rates Between Ages 25-34 From 2008 to 2025"
author: "Kei Matsumoto"
thanks: "Project repository available at: [https://github.com/keimatsumoto1/261A](https://github.com/keimatsumoto1/261A)"
date: 9/22/2025
date-format: long
abstract: "Employment through the years always changes with trends of industries with the rise of tech companies and now AI companies. However, with the increase in possible startup opportunities as well as rise in difficulty of job employment has unemployment rates changed over the years? Overall in this study, it is hard to find if the year correlates to the unemployment rates through linear regression as a different model is most likely better to analyze this data."
format: pdf
number-sections: true
---


```{r}
#| include: false
#| warning: false
#| message: false
# load libraries
library(readr)
library(ggplot2)
unemployment <- read_csv("unemployment_rate_by_age_groups.csv")
```

# Introduction

  Just by ear, there are people who say that there are better job opportunities with the rise in new advancements in AI and others stating that it is much more of a challenge than with the increase in people reaching for those positions. Most people will start their careers after graduating from either undergraduate or graduate school around the age of 25-34 which will be the age range that is analyzed. 
  
  This paper will only cover linear regression and linear modeling with a lack of further analysis. In this paper, we will conclude that linear regression may not be the optimal model to use to see the relationship between year and unemployment rates as the data highlights a possibility of a cyclic pattern. The linear regression will be analyzed through the programming language R using the basic functions within R and major packages such as “lm” and “ggplot” to visualize the data. 
  
  The paper will start with the description of the data used, into the methods conducted, and lastly the explanation of the results we have possibly found through the use of the statistical analysis within our knowledge of linear regression. 


# Data {#sec-data}

```{r fig-}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot comparing the year of survey study as the x axis, and the rate of unemployment between 0 and 1 as the y axis with the fitted line of regression."


ggplot(unemployment, aes(x = Year, y =  `Age 25-34`))+
  geom_point() +
  geom_smooth(method = "lm") + 
  xlab("Year") + 
  ylab("Unemployment Rate (0-1)")+
  ggtitle("Scatterplot of Relationship Between Year and Unemployment Rate Between Ages 25-34")+
  theme(plot.title = element_text(size = 12))


```

  This data was retrieved from the California Open Data Pool titled “Unemployment Rate by Age Groups” where the Current Population Survey (CPS) conducted a non-seasonally adjusted survey between 7 different age groups for each month of 2008 to 2025. The data is retrieved from the Employment Development Department who releases all of the employment statistics such as civilian labor forces, unemployment rates, and industry employment by geography. The EDD collaborates with the U.S. Bureau of Labor Statistics and the exact calculation of the unemployment rate is defined by (unemployed citizens / civilian labor force). Some months are missing in this dataset where January of 2008 and September through December of 2025 are not included. There does not appear to be any missing data within the dataset besides the last months of 2025. Data for 2025 may not be accurate as a quarter of the year is missing compared to other years due to the date of analysis is still during 2025. 

  The data uses measurements of percentages of unemployed people in the designated age group from the survey with other columns being demographic area, date, year, month, and each age group’s rate of unemployment.

# Methods {#sec-methods}

  In this study we will be conducting a simple linear regression to analyze the relationship between the rate of unemployment and the year.

$$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$

  $\beta_0$ will represent the intercept of the model which in this analysis will be the unemployment rate during the first time the study was conducted which is February of 2008.

  $\beta_1$ will represent the slope of the model which in this analysis will be the decrease in unemployment rate through each increase of 1 year.

  This model analysis will be conducted through the programming language R (insert citation) with several packages such as ggplot and readr.

# Results {#sec-results}

```{r lm}
#| echo: false
#| warning: false
#| message: false
lm = lm(`Age 25-34`~Year, data = unemployment)

```

Intercept = 6.56
Slope = -0.0032195
R^2 = 0.3847

  Within our simple linear regression we found that our  $\beta_0$is 6.56 and our $\beta_1$ is -0.0032195. Our $\beta_0$ in this scenario does not apply to our analysis due to the unemployment rate not being able to exceed 1. We can possibly say that our unemployment rate is close to 1 before 2008 which is due to the lack of studies made before 2008 in our experiment. Our $\beta_1$ is slightly negative indicating a very minor decrease over the years between 2008 and 2025. To be exact, for every additional year that passes, the rate of unemployment will decrease by 0.0032195 or 0.32195%. 

  In addition, our linear regression analysis includes and R-squared value of 0.3847 indicating that the fit of a linear model may not be ideal for this data. 


```{r residual}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Residual plot of the x axis being the fitted values and the y axis being the residuals for a simple linear regression with year being the respone and unemployment rate being the predictor."
p <- data.frame(
  res = resid(lm),
  fit = fitted(lm)
)
ggplot(p, aes(x = fit, y = res)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Residual (observed - fitted)") +
  xlab("Fitted value") +
  ggtitle("Residuals vs. Fitted plot")


```
```{r qq}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "QQ plot between the sample (x) and theoretical (y) quantiles"

residual = residuals(lm)
df1 = data.frame(residuals = residual)

qqnorm(residual)
qqline(residual)

```
  Our residuals vs fitted plots highlight that our model isn't the best fit for a simple linear regression where the variation of points are not in a form that shows no trends. With our data's sample size being 211 observations, we can possibly infer that our variances are equal.

  Our qq plot on the other hand suggests that our model has a possibility of having a good fit to our data. However, our tail and head of the QQ plot highlights the variance from the fit which can add on to our R-squared value being low.


# Discussion {#sec-discussion}

  In our analysis, we tested to see if we can fit a simple linear regression to find if unemployment rates have a relationship over time. We found that our data is not ideal to be analyzed through a simple linear regression compared to other models. Possible other models can seem to be a cyclical model because the scatter plot hints a trend in certain years. 2010 and 2020 have large peeks and sharp rises of unemployment rates which can be due to outside influences. One extreme hint for 2020 could be COVID 19 affecting the unemployment rate drastically. By noticing that the decrease then continues after the large spike, we can possibly see that if the unemployment rate would have slowly decreased over time without any interference. 


# References
H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.

R Core Team (2023). _R: A Language and Environment for Statistical Computing_. R Foundation for Statistical
  Computing, Vienna, Austria. <https://www.R-project.org/>.
  
Unemployment rate by age groups - Unemployment rate by age groups - California Open data. (n.d.).     
  https://data.ca.gov/dataset/unemployment-rate-by-age-groups/resource/be49fea7-af13-4781-8113-4bb66ba508e9
  
  
# Things that I will add in the next draft

1. Further depth in analysis and cleaning the wording
2. Citation with bib file
3. Citations within text
4. Possibly using more than just ages 25-34 and add a better combination graph with deeper analysis
5. sectioning
6. fix overall format with graphs taking up one page
  
