---
title: "Relationship Between Unemployment Rates and Time For People Ages 25-34 From 2008 to 2025"
author: "Kei Matsumoto"
thanks: "Project repository available at: [https://github.com/keimatsumoto1/261A](https://github.com/keimatsumoto1/261A)"
date: 9/22/2025
date-format: long
abstract: "Employment through the years always changes with new industry trends as well as international events/disasters. Job oppurtunities may have increased with the influence of new advancements such as A.I. but is this guaranteed to be an increase? Has unemployment rates changed over time? A dataset provided by  California Open Data highlights the overall unemployment rates over a monthly period from 2008 to 2025 within specific age groups. This paper will cover the possibility to fit a simple linear regression model to see if there is a relation between unemployment rate and time for people between ages 25 and 34. Overall in this study, it is difficult to conclude if time correlates to the unemployment rates through linear regression, as indications to fit a different model are evident."
format: pdf
number-sections: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false
# load libraries
library(readr)
library(ggplot2)
unemployment <- read_csv("unemployment_rate_by_age_groups.csv")
```

# Introduction

Just by ear, there are people who say that there are better job
opportunities with the rise in new advancements in A.I. On the other
hand, some others will say that it is much more of a challenge with the
increase in competitiveness for those positions. A recent study from
J.P. Morgan Global Research highlights how A.I. has both increased and
decreased employment rates depending on the work field. Areas such as
the white-collar sectors have received benefit through utilization of
A.I. while computer engineering college graduates are seeing
difficulties with the job market (J.P. Morgan Global Research, 2025). In
addition, a study conducted by the National Center for Education
Statistics (NCES) discovered that for 2019, many employment rates
increase with the higher degree of education attained (NCES, 2019). To
see the age groups of those who have finished at least a Bachelor's
Degree and possibly even a Master's Degree, we will research the age
group of 25-34 year old people.

This paper will only cover simple linear regression and linear modeling
with a lack of further analysis. In this paper, we will conclude that
linear regression may not be the optimal model to use to see the
relationship between time and unemployment rates as the data highlights
a possibility of a cyclic pattern due to sudden spikes of unemployment
in specific years. One large spike found was during 2019-2021 which can
be due to the breakout of Covid-19, with many jobs having in-person
interactions. The linear regression will be analyzed through the
programming language R using the basic functions within R and major
packages such as "readr" and "ggplot2" to initialize and visualize the
data.

The paper will start with the description of the data used, into the
methods conducted, the assumptions needed, and lastly the explanation of
the results we have possibly found through the use of the statistical
analysis within our knowledge of simple linear regression.

# Data {#sec-data}

```{r fig-}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot comparing the year of survey study as the x axis, and the rate of unemployment between 0 and 1 as the y axis with the fitted line of regression."


ggplot(unemployment, aes(x = Year, y =  `Age 25-34`))+
  geom_point() +
  geom_smooth(method = "lm") + 
  xlab("Year") + 
  ylab("Unemployment Rate (0-1)")+
  ggtitle("Scatterplot of Relationship Between Year and Unemployment Rate Between Ages 25-34")+
  theme(plot.title = element_text(size = 12))


```

This data was retrieved from California Open Data titled "Unemployment
Rate by Age Groups" where the Current Population Survey (CPS) conducted
a non-seasonally adjusted survey between 7 different age groups for each
month of 2008 to 2025. The data is retrieved from the Employment
Development Department who releases all of the employment statistics
such as civilian labor forces, unemployment rates, and industry
employment by geography (CPS, 2025). The EDD collaborates with the U.S.
Bureau of Labor Statistics and the exact calculation of the unemployment
rate is defined by (unemployed citizens / civilian labor force). Some
months are missing in this dataset where January of 2008 and September
through December of 2025 are not included. There does not appear to be
any missing data within the dataset besides the last months of 2025.
Therefore, the data for 2025 may not be as accurate as a quarter of the
year is missing compared to other years due to the date of analysis is
still being updated through each month.

In addition, we are only analyzing our data of people within the age of
25 to 34 years old. Our conclusions and analysis in this paper are
unable be used to predict the unemployment rates of other age groups as
they are completely different sources of data. Additional limitations
can be the reduced amount of data points for each month/year in
comparison to the entire dataset causing possible over fitting.

By analyzing the scatterplot above, we can notice that for each year
there are multiple points for every month. This can affect our variances
and violate homoscedacity with our unequal variance even if our number
of data points is high.

The data uses measurements of percentages of unemployed people in the
designated age group from the survey with other columns being
demographic area, date, year, month, and each age group's rate of
unemployment.

# Methods {#sec-methods}

In this study we will be conducting a simple linear regression to
analyze the relationship between the rate of unemployment and time
specifically in years ranging from 2008 to 2025.

$$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$

$\beta_0$ will represent the intercept of the model which in this
analysis will be the expected unemployment rate during the first time
the study was conducted which is February of 2008.

$\beta_1$ will represent the slope of the model which in this analysis
will be the expected change in unemployment rate through each increase
of 1 year.

$Y_i$ will represent the unemployment rate of the i-th Year from 2008 to
2025.

$X_i$ will represent the known constant which is the i-th year from 2008
to 2025.

$\epsilon_i$ will represent the random error with a mean of zero and
finite variance of $\sigma^2$.

This model analysis will be conducted through the programming language R
with several packages such as ggplot and readr (R Core Team, 2023).

# Assumptions

The assumptions we need to validate for our analysis are the following:

1.  Validity where the variables are appropriate for the question we are
    studying.
2.  Representativeness where the sample data represents the population
    of interest.
3.  Linearity where following function must be true:
    $E(Y_i | X_i) = \beta_0 + \beta_1X$
4.  Independence of errors where all errors of our model are assumed to
    be independent.
5.  Equal variance of errors where our errors' variances must be equal
    throughout the model.
6.  Normality of errors where our errors must be normally distributed.

# Results {#sec-results}

```{r lm}
#| echo: false
#| warning: false
#| message: false
lm = lm(`Age 25-34`~Year, data = unemployment)

```

By first analyzing the scatterplot made earlier, we can notice that the
trend is not in a linear formation with a trend with 2 different peaks
being evident. This shape will violate our assumption of linearity for
simple linear regression where our relationship between $Y_i$ and $X_i$
may not be linear and suggests a different possible model.

We also see that our data is dependent to time which reflects how our
observations are not independent. This will violate one of the key
assumptions (independent observations) to linear regression and suggest
that our standard errors are biased.

Within our simple linear regression, we found that our $\hat{\beta_0}$
is 6.56 and our $\hat{\beta_1}$ is -0.0032195. Our $\hat{\beta_0}$ in
this scenario does not apply to our analysis due to the unemployment
rate not being able to exceed 1. We can possibly say that our
unemployment rate is close to 1 before 2008 which is due to the lack of
studies made before 2008 in our experiment. Our $\hat{\beta_1}$ is
slightly negative indicating a very minor decrease over the years
between 2008 and 2025. To be exact, for every additional year that
passes, the rate of unemployment will on average decrease by 0.0032195
or 0.32195%. This indicates that may exist a very small but steady
decline in unemployment rates for those between 25 and 34.

In addition, we found that our R-squared value is 0.3847 indicating that
our model only explains around 38.47% of the total variance of our data.
This value is low emphasizing how our model may not be a significant fit
for the data.

```{r residual}
#| echo: false
#| fig-pos: "H"
#| warning: false
#| message: false
#| fig-cap: "Residual plot of the x axis being the fitted values and the y axis being the residuals for a simple linear regression with year being the response and unemployment rate being the predictor."
p <- data.frame(
  res = resid(lm),
  fit = fitted(lm)
)
ggplot(p, aes(x = fit, y = res)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Residual (observed - fitted)") +
  xlab("Fitted value") +
  ggtitle("Residuals vs. Fitted plot")


```

Our residuals vs fitted plots highlight that our model isn't the best
fit for a simple linear regression where the variation of points are not
in a form that shows no trends. This will violate multiple assumptions
such as error independence, error equal variance, and linearity. Even if
our data's sample size of 211 observations is quite large, these
assumptions are violated. This suggests possible transformations or
usage of a different model being highly recommended for this data.

```{r qq}
#| echo: false
#| fig-pos: "H"
#| warning: false
#| message: false
#| fig-cap: "QQ plot between the sample (x) and theoretical (y) quantiles with the line representing the theoretical normal distribution"

residual = residuals(lm)
df1 = data.frame(residuals = residual)

qqnorm(residual)
qqline(residual)

```

When observing our QQ plot at a glance, we can see that it might suggest
that our model has a possibility of having a good fit to our data.
However, our tail and head of the QQ plot highlights the variance from
the fit suggesting that our errors are not normal.

# Discussion {#sec-discussion}

In our analysis, we tested to see if we can fit a simple linear
regression to find if unemployment rates have a relationship with time.
We discovered that 2010 and 2020 have large peaks and sharp rises of
unemployment rates which can be due to outside influences.

2008 was one of our spikes in unemployment rates. The Great Recession
which was around the time can be a possible issue identified by the
large spike in the beginning of our data. Another spike was identified
around 2020 which highlights possibilities of influence by Covid-19
affecting the unemployment rate drastically. A study from US Bureau of
Statistics indicates that in 2019 found that roughly 3/4 of the workers
in the United States had to work with in-person human interaction (U.S.
Bureau of Labor Statistics, 2023). Combining this statistic with the
lock down protocol from Covid-19 can possibly connect to the drastic
increase in unemployment during 2020. By noticing that the decrease then
continues after the two large spikes, we can possibly see that the
unemployment rate possibly would have slowly decreased over time without
any interference.

During our analysis, we found that there has been multiple violations
against our assumptions with linear regression. These include
independent observations, linearity, error normality, error equal
variance, and error independence. These violations of assumptions
indicate that simple linear regression is not able to be used to analyze
the relation between time and unemployment rates.

Our scatterplots and our fitted vs observed residual plots indicate that
there are non linear trends. Due to the residual plot not having
randomness, our assumptions of error equal variance, normality, and
independence are violated. In addition, our R-squared value is low and
our QQ-plot also suggests that our errors are not considered normal.

Overall, we can conclude that a different or transformation can be
possibly used to fit this data. We discovered that there may exist a
small negative correlation between time and unemployment rates but we
are unable to justify this through simple linear regression. Our data is
found to be dependent to time with indications of a cyclical pattern
from observing the scatter plot.

# References

H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag
New York, 2016.

Morgan, J. P. (2025). AI's Impact on Job Growth \| J.P. Morgan Global
Research. Retrieved from Jpmorgan.com website:
https://www.jpmorgan.com/insights/global-research/artificial-intelligence/ai-impact-job-growth

R Core Team (2023). *R: A Language and Environment for Statistical
Computing*. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>.

Relationship Between Educational Attainment and Labor Underutilization.
(n.d.). Retrieved from nces.ed.gov website:
https://nces.ed.gov/pubs2019/2019039/index.asp

Unemployment rate by age groups - Unemployment rate by age groups -
California Open data. (n.d.).\
https://data.ca.gov/dataset/unemployment-rate-by-age-groups/resource/be49fea7-af13-4781-8113-4bb66ba508e9

U.S. Bureau of Labor Statistics. (n.d.). Three-fourths of workers had to
interact with the public in 2019; 4.3 percent worked around crowds. U.S.
Bureau of Labor Statistics.
https://www.bls.gov/opub/ted/2020/three-fourths-of-workers-had-to-interact-with-the-public-in-2019-4-3-percent-worked-around-crowds

Wickham H, Hester J, Bryan J (2023). \_readr: Read Rectangular Text
Data\_. R package version 2.1.4,
https://CRAN.R-project.org/package=readr.
